---
title: "Data_preprocessing"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r echo = FALSE, message=FALSE}
# Carregar llibreries
library(car)
library(tables)   # tabular
library(emmeans)  # emmeans
library(car)      # residualPlot
library(RcmdrMisc)#
library(multcomp) # cld 
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)  #
```

```{r echo = FALSE, message=FALSE}
dd    <- read.csv("C:/Users/javil/OneDrive/Documentos/UPC/2n/ME/Proyecto/demographic/demographic.csv") # Llegir
summary(dd)
```
```{r echo = FALSE, message=FALSE}


dd$DMDBORN4 <- as.numeric(dd$DMDBORN4)
dd$DMDBORN4[dd$DMDBORN4 %in% c(77,99) | is.na(dd$DMDBORN4)] <- 0
# DMDCITZN: valores existentes [1,2, NA, 7,9]
dd$DMDCITZN <- as.numeric(dd$DMDCITZN)
dd$DMDCITZN[dd$DMDCITZN %in% c(7,9) | is.na(dd$DMDCITZN)] <- 0

dd$DMDEDUC2 <- as.numeric(dd$DMDEDUC2)
dd$DMDEDUC2[dd$DMDEDUC2 %in% c(7,9) | is.na(dd$DMDEDUC2)] <- 0

# DMDHRBR4: valores existentes [1,2, NA,77]
dd$DMDHRBR4 <- as.numeric(dd$DMDHRBR4)
dd$DMDHRBR4[dd$DMDHRBR4 %in% c(77) | is.na(dd$DMDHRBR4)] <- 0

# DMDHREDU: valores existentes [1,2,3,4,5,NA,7,9]
dd$DMDHREDU <- as.numeric(dd$DMDHREDU)
dd$DMDHREDU[dd$DMDHREDU %in% c(7,9) | is.na(dd$DMDHREDU)] <- 0 

dd$DMQMILIZ <- as.numeric(dd$DMQMILIZ)
dd$DMQMILIZ[dd$DMQMILIZ %in% c(7,9) | is.na(dd$DMQMILIZ)] <- 0 

# DMDHRMAR: valores existentes [1,2,3,4,5,6,NA,77,99]
dd$DMDHRMAR <- as.numeric(dd$DMDHRMAR)
dd$DMDHRMAR[dd$DMDHRMAR %in% c(77,99) | is.na(dd$DMDHRMAR)] <- 0 



# 2. DEFINIR VARIABLES CATEGÓRICAS
categorical_vars <- c(
  "AIALANGA", "FIALANG", "MIALANG", "SIALANG",
  "FIAINTRP", "MIAINTRP", "SIAINTRP",
  "FIAPROXY", "MIAPROXY", "SIAPROXY",
  "DMDBORN4", "DMDHRBR4", "DMDCITZN",
  "DMDEDUC2", "DMDEDUC3", "DMDHREDU", "DMDHSEDU",
  "RIAGENDR", "DMDHRGND",
  "DMDMARTL", "DMDHRMAR",
  "RIDRETH1", "RIDRETH3",
  "DMQMILIZ", "DMQADFC",
  "RIDEXPRG", "RIDEXMON", "RIDSTATR", "SDDSRVYR",
  "INDFMIN2", "INDHHIN2"
)

# Convertir a factores 
categorical_vars_existentes <- categorical_vars[categorical_vars %in% names(dd)]
dd <- dd %>%
  mutate(across(all_of(categorical_vars_existentes), as.factor))

summary(dd)

num_vars <- names(dd)[sapply(dd, is.numeric)]

for (var in num_vars) {
  x <- dd[[var]]
  hist(
    x[!is.na(x)],
    main = paste("Distribución de", var, "(sin NA)"),
    xlab = var,
    ylab = "Frecuencia",
    col = "lightblue",
    border = "white"
  )
}

# --- CATEGÓRICAS ---
cat_vars <- names(dd)[sapply(dd, function(x) is.factor(x) || is.character(x))]

for (var in cat_vars) {
  x <- dd[[var]]
  # Eliminar NA y convertir a factor
  x <- factor(x[!is.na(x)])
  
  barplot(
    table(x),
    main = paste("Distribución de", var, "(sin NA)"),
    xlab = var,
    ylab = "Frecuencia",
    col = "lightgreen",
    border = "white",
    las = 2,          # rota etiquetas del eje x si son largas
    cex.names = 0.8   # ajusta tamaño de las etiquetas
  )
}
```

```{r echo = FALSE, message=FALSE}

install.packages("tidyverse")
library(tidyverse)

dd_clean <- dd

resultados <- data.frame(variable = character(), 
                         asociacion = numeric(), 
                         tipo = character(),
                         stringsAsFactors = FALSE)

# A) Variables NUMÉRICAS
vars_numericas <- dd_clean %>% 
  select(where(is.numeric), -INDFMPIR) %>% 
  names()

for(var in vars_numericas) {
  if(var %in% names(dd_clean)) {
    correlacion <- cor(dd_clean[[var]], dd_clean$INDFMPIR, 
                       use = "pairwise.complete.obs")
    if(!is.na(correlacion)) {
      resultados <- rbind(resultados, 
                         data.frame(variable = var, 
                                   asociacion = correlacion,
                                   tipo = "Numérica (r)"))
    }
  }
}
# B) Variables CATEGÓRICAS - Eta (Correlation Ratio)
vars_categoricas <- dd_clean %>% 
  select(where(is.factor)) %>% 
  names()

for(var in vars_categoricas) {
  if(var %in% names(dd_clean) && sum(!is.na(dd_clean[[var]])) > 0) {
    formula <- as.formula(paste("INDFMPIR ~", var))
    
    tryCatch({
      modelo <- aov(formula, data = dd_clean)
      ss <- summary(modelo)[[1]]$"Sum Sq"
      
      eta_squared <- ss[1] / sum(ss)
      eta <- sqrt(eta_squared)
      
      resultados <- rbind(resultados, 
                         data.frame(variable = var, 
                                   asociacion = eta,
                                   tipo = "Categórica (η)"))
    }, error = function(e) {
      # Si falla, omitir
    })
  }
}

# Ordenar resultados
resultados <- resultados %>%
  arrange(desc(abs(asociacion))) %>%
  mutate(
    variable_label = paste0(variable, " (", tipo, ")"),
    direccion = ifelse(tipo == "Numérica (r)" & asociacion < 0, 
                       "Negativa", "Positiva")
  )

# Ver tabla completa
print(resultados)

top_15 <- resultados %>%
  arrange(desc(abs(asociacion))) %>%
  head(15)

grafica1 <- ggplot(top_15, aes(x = reorder(variable, abs(asociacion)), 
                                y = asociacion,
                                fill = tipo)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.3f", asociacion)), 
            hjust = ifelse(top_15$asociacion > 0, -0.1, 1.1),
            size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("Numérica (r)" = "#2c7bb6", 
                                "Categórica (η)" = "#d7191c")) +
  theme_minimal(base_size = 12) +
  labs(title = "TOP 15: Variables más asociadas con INDFMPIR",
       subtitle = "Correlación de Pearson (r) para numéricas | Correlation Ratio (η) para categóricas",
       x = "Variable",
       y = "Fuerza de asociación",
       fill = "Tipo") +
  theme(axis.text.y = element_text(size = 11, face = "bold"),
        plot.title = element_text(size = 14, face = "bold"),
        legend.position = "bottom")

print(grafica1)

medio_15 <- resultados %>%
  arrange(desc(abs(asociacion))) %>%
  slice(16:30)

if(nrow(medio_15) > 0) {
  grafica2 <- ggplot(medio_15, aes(x = reorder(variable, abs(asociacion)), 
                                    y = asociacion,
                                    fill = tipo)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = sprintf("%.3f", asociacion)), 
              hjust = ifelse(medio_15$asociacion > 0, -0.1, 1.1),
              size = 4, fontface = "bold") +
    coord_flip() +
    scale_fill_manual(values = c("Numérica (r)" = "#2c7bb6", 
                                  "Categórica (η)" = "#d7191c")) +
    theme_minimal(base_size = 12) +
    labs(title = "Posiciones 16-30: Variables con asociación media",
         subtitle = "Correlación de Pearson (r) para numéricas | Correlation Ratio (η) para categóricas",
         x = "Variable",
         y = "Fuerza de asociación",
         fill = "Tipo") +
    theme(axis.text.y = element_text(size = 11, face = "bold"),
          plot.title = element_text(size = 14, face = "bold"),
          legend.position = "bottom")
  
  print(grafica2)
}

resto <- resultados %>%
  arrange(desc(abs(asociacion))) %>%
  slice(31:n())

if(nrow(resto) > 0) {
  grafica3 <- ggplot(resto, aes(x = reorder(variable, abs(asociacion)), 
                                 y = asociacion,
                                 fill = tipo)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = sprintf("%.3f", asociacion)), 
              hjust = ifelse(resto$asociacion > 0, -0.1, 1.1),
              size = 3.5) +
    coord_flip() +
    scale_fill_manual(values = c("Numérica (r)" = "#2c7bb6", 
                                  "Categórica (η)" = "#d7191c")) +
    theme_minimal(base_size = 11) +
    labs(title = "Resto de variables: Asociación débil",
         subtitle = "Correlación de Pearson (r) para numéricas | Correlation Ratio (η) para categóricas",
         x = "Variable",
         y = "Fuerza de asociación",
         fill = "Tipo") +
    theme(axis.text.y = element_text(size = 10),
          plot.title = element_text(size = 14, face = "bold"),
          legend.position = "bottom")
  
  print(grafica3)
}


top_20_heatmap <- resultados %>%
  arrange(desc(abs(asociacion))) %>%
  head(20)

heatmap <- ggplot(top_20_heatmap, aes(x = 1, y = reorder(variable, abs(asociacion)), 
                                       fill = asociacion)) +
  geom_tile(color = "white", size = 1.5) +
  geom_text(aes(label = sprintf("%.3f", asociacion)), 
            color = "black", fontface = "bold", size = 5) +
  scale_fill_gradient2(low = "#d7191c", mid = "#ffffbf", high = "#2c7bb6",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal(base_size = 12) +
  labs(title = "HEATMAP: Top 20 Variables más asociadas con INDFMPIR",
       x = "", y = "",
       fill = "Asociación") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        axis.text.y = element_text(size = 11, face = "bold"),
        plot.title = element_text(size = 14, face = "bold"))

print(heatmap)


```
#Description
'RIAGENDR',          # género del participante, factor, 1 hombre 2 mujer
    'RIDAGEYR',          # edad del participante, numerica
    'RIDRETH3',          # raza / etnia, factor 1:Mexican American 2:Other Hispanic 3:Non-Hispanic White 4:Non-Hispanic Black 6:Non-Hispanic Asian 7:Other Race - Including Multi-Racial
    'DMDBORN4',          # país de nacimiento, factor 1:Born in 50 US states or Washington, DC 2:others 77:Refused 99: Don't know
    'DMDCITZN',          # ciudadanía, factor
    'SIALANG',           # idioma de entrevista, factor, 1 english 2 spanish
    'SIAINTRP',          # uso de intérprete, factor 1 yes 2 no
    'DMDFMSIZ',          # tamaño de la familia, numerica discreta
    'DMDHHSZA',          # niños <5, factor 0:none 1:1, 2:2, 3:3 o mas
    'DMDHHSZB',          # niños 6–17, factor 0:none 1:1, 2:2, 3:3, 4:4 o mas
    'DMDHHSZE',          # adultos ≥60, factor, 0:none 1:1, 2:2, 3:3 o mas
    'DMDHRGND',          # género del cabeza de familia, factor 1 hombre 2 mujer
    'DMDHRAGE',          # edad del cabeza de familia, numerica discreta
    'DMDHRBR4',          # país de nacimiento del cabeza, factor 1:Born in 50 US states or Washington, DC 2:others 77:Refused 99: Don't know
    'DMDHREDU',          # educación del cabeza, factor 1:Less than 9th grade 2:9-11th grade (Includes 12th grade with no diploma) 3:High school graduate/GED or equivalent 4:Some college or AA degree 5:College graduate or above 7:refused 9:Don't know
    'DMDHRMAR'           # estado civil del cabeza, factor 1:Married 2:Widowed 3:Divorced 4:Separated 5:Never married 6:Living with partner 77:Refused 99:Don't know

```{r echo = FALSE, message=FALSE}
selected_features <- c(
  'RIAGENDR', 'RIDAGEYR', 'DMDEDUC2', 'RIDRETH3', 'DMDBORN4', 'DMDCITZN',
  'SIALANG', 'SIAINTRP', 'DMDFMSIZ', 'DMDHHSZA', 'DMDHHSZB',
  'DMDHHSZE', 'DMDHRGND', 'DMDHRAGE', 'DMDHRBR4', 'DMDHREDU',
  'DMDHRMAR', 'WTINT2YR','DMQMILIZ', 'INDFMPIR' 
)
dd <- dd[, colnames(dd) %in% selected_features]
summary(dd)
```
#there are missing values in target variable
```{r echo = FALSE, message=FALSE}
dd$INDFMPIR[dd$INDFMPIR <= 0] <- NA
dd[dd$INDFMPIR <= 0, ]
dd$INDFMPIR[dd$INDFMPIR <= 0] <- NA
dd$INDFMPIR[is.na(dd$INDFMPIR)] <- median(dd$INDFMPIR, na.rm = TRUE)

dd_binary <- dd %>%
  filter(!is.na(DMQMILIZ), DMQMILIZ != 0) %>%   # elimina NA y 0
  mutate(DMQMILIZ = ifelse(DMQMILIZ == 2, 0, 1))
#there are values that are 0 in the codification, in the NHANES documentation 0 is said to be as missing values, or could not be determined
#we will assume it as missing data

summary(dd_binary)
```

```{r echo = FALSE, message=FALSE}

par(mfrow=c(2,5), mar=c(2,2,2,1)) 
colors <- c(2,3)                 
group_var <- "DMQMILIZ"   

vars <- setdiff(selected_features, "INDFMPIR")

vars <- setdiff(vars, group_var)

group_levels <- unique(na.omit(dd_binary[[group_var]]))
if(length(group_levels) == 2){
  colors <- c("tomato","seagreen")
} else {
  colors <- grDevices::rainbow(length(group_levels))
}

# -------------- Preparar layout --------------
# dimensiones automáticas para que no se corte: ncol por defecto 4
n_vars <- length(vars)
ncol_plot <- 4
nrow_plot <- ceiling(n_vars / ncol_plot)
oldpar <- par(no.readonly = TRUE)
par(mfrow = c(nrow_plot, ncol_plot), mar = c(3,3,3,1), oma = c(0,0,2,0))

# Boxplot if numeric and barplot if categorical
for (v in vars) {
  # si variable NO es factor (numérica) -> boxplot por grupo (horizontal)
  if (is.numeric(dd_binary[[v]])) {
    # Convertir grupo a factor para orden consistente en boxplot
    grp <- factor(dd_binary[[group_var]])
    form <- as.formula(paste(v, "~", group_var))
    # boxplot con colores por nivel del grupo
    boxplot(form, data = dd_binary, main = v, col = colors, horizontal = TRUE, outline = TRUE,
            xlab = "", ylab = "", las = 1)
  } else {
    # variable categórica -> barplot de proporciones por columna (similar ejemplo)
    # tabla de contingencia: rows = group, cols = categories of v
    tab <- table(dd_binary[[group_var]], dd_binary[[v]], useNA = "ifany")
    # prop.table por columnas (proporción dentro de cada categoría de v)
    prop_tab <- prop.table(tab, margin = 2)
    # barplot (stacked) con leyenda
    barplot(prop_tab, main = v, col = colors, beside = FALSE, las = 1)
    # añadir leyenda si hay espacio
    if (length(group_levels) <= 6) {
      legend("topright", legend = levels(factor(dd_binary[[group_var]])), fill = colors, cex = 0.8, bty = "n")
    }
  }
}

hist(
  dd$INDFMPIR[!is.na(dd$INDFMPIR)],
  main = "Distribución de INDFMPIR (sin NA)",
  xlab = "INDFMPIR",
  ylab = "Frecuencia",
  col = "lightblue",
  border = "white"
)
```

#the distribution is not normal, as the values are a small percentage of the data we could impute by the median

```{r echo = FALSE, message=FALSE}

par(mfrow=c(1,2))
with(dd, plotMeans(INDFMPIR, SIALANG, SIAINTRP, error.bars="conf.int", level=0.95, connect=TRUE))
with(dd, plotMeans(INDFMPIR, SIAINTRP, SIALANG, error.bars="conf.int", level=0.95, connect=TRUE))

library(ggplot2)

ggplot(dd, aes(x = DMDEDUC2, y = INDFMPIR, fill = RIAGENDR)) +
  geom_boxplot() +
  labs(title = "Interacción: Educación × Género sobre INDFMPIR",
       x = "Nivel educativo",
       y = "INDFMPIR") +
  theme_minimal()
#seems to be a little interaction

ggplot(dd, aes(x = DMDBORN4, y = INDFMPIR, fill = SIALANG)) +
  geom_boxplot() +
  labs(title = "Interacción: born x language",
       x = "Nivel educativo",
       y = "INDFMPIR") +
  theme_minimal()
#not much

ggplot(dd, aes(x = DMDEDUC2, y = INDFMPIR, fill = SIALANG)) +
  geom_boxplot() +
  labs(title = "Interacción: education x language",
       x = "Nivel educativo",
       y = "INDFMPIR") +
  theme_minimal()

ggplot(dd, aes(x = DMDBORN4, y = INDFMPIR, fill = DMDCITZN)) +
  geom_boxplot() +
  labs(title = "Interacción: born x citzn",
       x = "Nivel educativo",
       y = "INDFMPIR") +
  theme_minimal()
#not much
ggplot(dd, aes(x = SIAINTRP, y = INDFMPIR, fill = DMDCITZN)) +
  geom_boxplot() +
  labs(title = "Interacción: intrp x citzn",
       x = "Nivel educativo",
       y = "INDFMPIR") +
  theme_minimal()
#not much

#we have tried modified models with each interaction, the only relevaant for model that we should include is dmdborn4 x sialang
```
#gaussian and identity link is not a good option but we do it to test
```{r echo=FALSE, message=FALSE}

mlgz <- glm(INDFMPIR ~ ., dd, family = gaussian(link=identity))
summary(mlgz)

Anova(mlgz, test.statistic = 'LR')

#deviance not good



```

```{r echo=FALSE, message=FALSE}
residualPlots(mlgz, layout = c(2,1), ask = FALSE, tests = FALSE)
```
#we can see the line of the model is not good at all, its what we expected

```{r echo=FALSE, message=FALSE}
influencePlot(model = mlgz)
```

#330 is an important outlier
```{r echo=FALSE, message=FALSE}
dd[330, ]
dd[1868, ]
dd[213, ]

```
#the outliers seem to be relevant 

#lets change the family and link function
```{r echo = FALSE, message=FALSE}

mlgz2 <- glm(INDFMPIR ~ .+ DMDEDUC2*SIALANG, dd, family = gaussian(link=log))
summary(mlgz2)

Anova(mlgz2, test.statistic = 'LR')

residualPlots(mlgz2, layout = c(2,1), ask = FALSE, tests = FALSE)

```


```{r echo = FALSE, message=FALSE}

mlgz3 <- glm(INDFMPIR ~ . + DMDEDUC2*SIALANG, dd, family = Gamma(link=log))
summary(mlgz3)

Anova(mlgz3, test.statistic = 'LR')

residualPlots(mlgz3, layout = c(2,1), ask = FALSE, tests = FALSE)

vif(mlgz3) 



```


```{r echo = FALSE, message=FALSE}
influencePlot(model=mlgz3)
```



```{r echo = FALSE, message=FALSE}
# it has so much missing values, but we cant work on the rows without NA
table(dd_binary$DMQMILIZ)

# Separar por clase
casos_0 <- which(dd_binary$DMQMILIZ == 0)
casos_1 <- which(dd_binary$DMQMILIZ == 1)

# Número de casos a mantener de la clase 0
n_0_nuevo <- length(casos_1)

# Seleccionar aleatoriamente n_0_nuevo casos de los 0's
set.seed(123)  # para reproducibilidad
casos_0_muestreados <- sample(casos_0, size = n_0_nuevo, replace = FALSE)

# Combinar los índices de las dos clases
indices_finales <- sort(c(casos_0_muestreados, casos_1))

# Crear dataset balanceado
dd_binary_balanced <- dd_binary[indices_finales, ]

levels(factor(dd_binary_balanced$DMDBORN4))
levels(factor(dd_binary$DMDBORN4))

# Verificar el balance
table(dd_binary_balanced$DMQMILIZ)
table(dd_binary$DMQMILIZ)

vars_categoricas <- c("RIAGENDR", "RIDRETH3", "DMQMILIZ", "DMDBORN4", "DMDCITZN", 
                      "DMDEDUC2", "SIALANG", "SIAINTRP", "DMDHRGND", "DMDHRBR4", "DMDHREDU", "DMDHRMAR")

# Threshold: elimina categorías con menos de N observaciones
threshold <- 200 # Cambia este número según lo que consideres "raro"

# Para cada variable categórica, elimina filas con categorías raras
for (var in vars_categoricas) {
  if (var %in% names(dd_binary)) {
    # Contar frecuencias
    freq <- table(dd_binary[[var]], useNA = "ifany")
    
    # Identificar categorías raras (menos de threshold)
    categorias_raras <- names(freq[freq < threshold])
    
    if (length(categorias_raras) > 0) {
      cat(var, "- Categorías raras (< ", threshold, "):", categorias_raras, "\n")
      
      # Eliminar filas con categorías raras
      dd_binary <- dd_binary[!(dd_binary[[var]] %in% categorias_raras), ]
    }
  }
}


summary(dd_binary)

summary(mb_balanced <- glm(DMQMILIZ ~ ., dd_binary_balanced, family=binomial(link='logit')))

Anova(mb_balanced, test.statistic = 'LR')

pred_balanced <- predict(mb_balanced, newdata = dd_binary, type = "response")
roc(dd_binary$DMQMILIZ, pred_balanced)


summary(mb <- glm(DMQMILIZ ~ ., dd_binary, family=binomial(link='logit')))
#we cann clearly see a case of quasi-separation on this model, solutions --> reduce complexity or use penalised regression

Anova(mb, test.statistic = 'LR')
#lets remove columns that have p-value insignificant

residualPlots(mb, layout = c(2,1), ask = FALSE, tests = FALSE)

pred <- predict(mb, type = "response")
roc(dd_binary$DMQMILIZ, pred)

summary(mb2 <- glm(DMQMILIZ ~ RIAGENDR + RIDAGEYR + RIDRETH3 + DMDBORN4 + DMDCITZN + DMDEDUC2 + SIALANG + DMDHHSZA + WTINT2YR, dd_binary, family=binomial(link='logit')))

Anova(mb2, test.statistic = 'LR')

residualPlots(mb2, layout = c(2,1), ask = FALSE, tests = FALSE)

pred2 <- predict(mb2, type = "response")
roc(dd_binary$DMQMILIZ, pred2)

#lets try penalised firth regression to solve this problem

library(logistf)
install.packages("pROC")
library(pROC)

mb2_firth <- logistf(
  DMQMILIZ ~ RIAGENDR + RIDAGEYR + RIDRETH3 + DMDBORN4 +
    DMDCITZN + DMDEDUC2 + SIALANG + DMDHHSZA + WTINT2YR,
  data = dd_binary
)
summary(mb2_firth)


pred <- predict(mb2_firth, type = "response")
roc(dd_binary$DMQMILIZ, pred)



#area: 0.912 


```



